---
title: "Data on the Rocks: Touring U.S. Lighthouses"
name: Verity Elliott
output: html_document
---

Lighthouses are singular pieces of coastal infrastructure that serve a functional purpose - guiding mariners - but also embody geographic and historical narratives. The Wikipedia page “List of lighthouses in the United States” compiles standardized information on U.S. lighthouses including location (latitude/longitude), year established, tower height, range, and notes on status. Rather than manually copying entries or using a secondary dataset, we scrape the canonical list directly from Wikipedia. This ensures reproducibility and ties our analysis to a visible public source that encodes the selection and categorization decisions of the community that maintains the page.

```{r load-packages, warning = FALSE, message=FALSE}
# Load all required packages quietly
suppressMessages({
  library(tidyverse)
  library(rvest)
  library(httr)
  library(ggrepel)
  library(viridis)
})
```

The lighthouse data are not stored in a single master table on Wikipedia. Instead, the page Lists of lighthouses in the United States serves as an index linking to individual state-level pages. Each state page contains one or more tables listing lighthouses and their attributes. To build a comprehensive dataset, we first retrieve the index page, extract all links to state-specific lighthouse pages, and then programmatically iterate over those links to collect their tables.

```{r retrieve_data}
# ---- 1. Read the index page (main lighthouse list page) ----
index_url <- "https://en.wikipedia.org/wiki/List_of_lighthouses_in_the_United_States"

index_page <- GET(
  index_url,
  user_agent("Mozilla/5.0 (R lighthouse data project)")
) %>% read_html()

# ---- 2. Extract links to state-level lighthouse pages ----
state_links <- index_page %>%
  html_elements("a") %>%
  html_attr("href") %>%
  na.omit() %>%
  unique() %>%
  # Only keep links starting with lighthouse list pattern
  keep(~ str_detect(., "^/wiki/List_of_lighthouses_in_")) %>%
  # Add base URL
  paste0("https://en.wikipedia.org", .)

# Remove main U.S. page and known non-state pages
state_links <- state_links %>%
  keep(~ !str_detect(.x, "List_of_lighthouses_in_the_United_States$") &
         !str_detect(.x, "List_of_lighthouses_in_the_United_States_territories")) 

length(state_links)  # Should be roughly 50+

# ---- 3. Function to retrieve lighthouse tables from a single state page ----
get_lighthouse_tables <- function(url) {
  tryCatch({
    page <- read_html(url)
    
    # Clean state name from URL
    state_name <- url %>%
      str_replace("https://en.wikipedia.org/wiki/List_of_lighthouses_in_", "") %>%
      str_replace_all("_", " ") %>%
      str_replace_all("\\s*\\(.*\\)", "") %>%  # remove parentheses
      str_replace_all("%[0-9A-F]{2}", "") %>% # rough URL decoding
      str_replace_all("#.*$", "") %>%         # remove fragments like #Illinois
      str_trim()
    
    # Extract tables with class wikitable
    tables <- page %>% html_elements("table.wikitable")  
    tables <- map(tables, function(tab) {
      df <- html_table(tab, fill = TRUE)
      
      # Only keep tables with a Name column
      if (!any(str_detect(names(df), regex("name", ignore_case = TRUE)))) return(NULL)
      
      # Standardize column names
      names(df) <- names(df) %>%
        {ifelse(is.na(.) | . == "", paste0("V", seq_along(.)), .)} %>%
        tolower() %>%
        str_replace_all("\\s+", "_")
      
      # Add state column
      df$state <- state_name
      
      # Extract numeric height from any column containing "height"
      height_col <- grep("height", names(df), ignore.case = TRUE)
      if (length(height_col) > 0) {
        df$height <- df[[height_col[1]]]  # take first matching height column
      } else {
        df$height <- NA_character_
      }
      
      df
    })
    
    tables <- compact(tables)  # remove NULL tables
    tables
    
  }, error = function(e) {
    message("Failed to read ", url, ": ", e$message)
    NULL
  })
}

# ---- 4. Scrape all state pages with a short pause between requests ----
all_tables <- map(state_links, ~{
  Sys.sleep(0.5)  # polite delay
  get_lighthouse_tables(.x)
})

# ---- 5. Flatten list of tables ----
all_tables <- flatten(all_tables)

# ---- 6. Standardize column names and convert all columns to character ----
all_tables <- map(all_tables, ~{
  df <- .
  
  colnames(df) <- colnames(df) %>%
    {ifelse(is.na(.) | . == "", paste0("V", seq_along(.)), .)} %>%
    tolower() %>%
    str_replace_all("\\s+", "_")
  
  # Ensure essential columns exist
  if(!"state" %in% colnames(df)) df$state <- NA_character_
  if(!"name" %in% colnames(df)) df$name <- NA_character_
  if(!"location" %in% colnames(df)) df$location <- NA_character_
  if(!"height" %in% colnames(df)) df$height <- NA_character_
  
  colnames(df) <- make.unique(colnames(df), sep = "_dup_")
  df %>% mutate(across(everything(), as.character))
})

# ---- 7. Combine into master dataframe ----
lighthouses_all <- bind_rows(all_tables)

# ---- 8. Clean and select relevant columns ----
lighthouses_clean <- lighthouses_all %>%
  fill(state, .direction = "down") %>%
  mutate(
    across(where(is.character), ~ str_squish(.) %>% str_replace_all("\n", " ")),
    year_first_lit = str_extract(year_first_lit, "\\d{4}") %>% as.integer(),
    height_ft = str_extract(height, "\\d+\\.?\\d*") %>% as.numeric()
  ) %>%
  filter(!is.na(name) & name != "") %>%
  select(-height)  # keep numeric height_ft only

# ---- 9. Sanity Checks ----
dim(lighthouses_clean)
head(lighthouses_clean, 10)
```

Although we previously standardized many fields, state tables still contain column-name variants (e.g., multiple height or automation columns). In this step, we consolidate these variants, standardize naming conventions, and finalize the structure of the master dataset.

```{r convert_data}
lighthouses_clean <- lighthouses_all %>%
  # ---- 1. Ensure necessary columns exist ----
  mutate(
    # Ensure height variants
    height = if ("height" %in% colnames(.)) height else NA_character_,
    height_ft = if ("height_ft" %in% colnames(.)) height_ft else NA_character_,
    
    # Ensure automated variants exist
    automated = if ("automated" %in% colnames(.)) automated else NA_character_,
    year_automated = if ("year_automated" %in% colnames(.)) year_automated else NA_character_
  ) %>%
  
  # ---- 2. Select relevant columns plus potential variants ----
  select(any_of(c(
    "state",
    "name",
    "location",
    "coordinates",
    "year_first_lit",
    "automated", "year_automated",
    "current_lens",
    "height", "height_ft"
  ))) %>%
  
  # ---- 3. Fill down state values ----
  fill(state, .direction = "down") %>%
  
  # ---- 4. Clean text fields ----
  mutate(across(where(is.character), ~ str_squish(.) %>% str_replace_all("\n", " "))) %>%
  
  # ---- 5. Consolidate height columns to a single numeric column ----
  mutate(
    height = coalesce(height, height_ft),
    height_ft = str_extract(height, "\\d+\\.?\\d*") %>% as.numeric()
  ) %>%
  
  # ---- 6. Consolidate automated column safely ----
  mutate(
    automated = coalesce(automated, year_automated)
  ) %>%
  
  # ---- 7. Remove rows without a lighthouse name ----
  filter(!is.na(name) & name != "") %>%
  
  # ---- 8. Drop temporary or redundant columns ----
  select(-c(height, year_automated)) %>%
  
  # ---- 9. Reorder columns exactly as desired ----
  select(
    state,
    name,
    location,
    coordinates,
    year_first_lit,
    automated,
    current_lens,
    height_ft
  )

# ---- Sanity Checks ----
# Check number of rows and columns
dim(lighthouses_clean)

# Preview first 10 rows of key columns
key_cols <- c("name", "location", "coordinates", "year_first_lit",
              "automated", "current_lens", "height_ft", "state")
head(lighthouses_clean %>% select(any_of(key_cols)), 10)

# Inspect structure
glimpse(lighthouses_clean)
```

Many of the coordinates in the dataset are currently stored as crude strings, often in degrees with directional indicators (e.g., "42.9672°N 70.6231°W"). To make these usable for mapping and spatial analysis, we convert the coordinates into decimal degrees. This involves splitting the latitude and longitude, removing any directional symbols, and applying the correct sign based on whether the coordinate is north/south or east/west. After this transformation, the dataset will have numeric lat and lon columns, allowing us to plot lighthouse locations on maps or perform geographic filtering.

```{r convert_coordinates}
lighthouses_clean <- lighthouses_clean %>%
  mutate(
    # Remove parentheses/brackets and extra spaces
    coords_clean = str_remove_all(coordinates, "\\(.*?\\)|\\[.*?\\]") %>% str_squish(),

    # Extract lat/lon with decimal and hemisphere
    lat_raw = str_extract(coords_clean, "[-]?\\d+\\.?\\d*°?\\s*[NS]"),
    lon_raw = str_extract(coords_clean, "[-]?\\d+\\.?\\d*°?\\s*[EW]"),

    # Convert to numeric
    lat = as.numeric(str_remove_all(lat_raw, "[°NS]")),
    lon = as.numeric(str_remove_all(lon_raw, "[°EW]")),

    # Apply sign based on hemisphere
    lat = ifelse(str_detect(lat_raw, "S"), -lat, lat),
    lon = ifelse(str_detect(lon_raw, "W"), -lon, lon)
  ) %>%
  filter(!is.na(lat) & !is.na(lon))

# ---- Sanity Check ----
# Preview first 10 rows of lat/lon
lighthouses_clean %>% select(name, state, lat, lon) %>% head(10)
```

### Lighthouse Lens: Explore the Data

Once the lighthouse dataset has been cleaned and coordinates converted, it becomes useful to explore it dynamically. A simple way to do this is by defining a search function that allows filtering by one or more criteria such as state, name, age, or height. This enables quick querying of the dataset to answer questions like “Which lighthouses are taller than 150 feet?” or “Show all lighthouses built before 1850 in Massachusetts.” By wrapping these filters into a function, the dataset becomes interactive and easier to analyze or visualize on a map.

```{r search_function}
# Function to search/filter lighthouses
search_lighthouses <- function(df,
                               name_search = NULL,
                               state_search = NULL,
                               location_search = NULL,
                               coordinates_search = NULL,
                               min_year = NULL,
                               max_year = NULL,
                               automated_search = NULL,
                               lens_search = NULL,
                               min_height = NULL,
                               max_height = NULL) {
  
  results <- df
  
  # Helper for safe text filtering
  safe_text_filter <- function(column, value) {
    str_detect(column, fixed(value, ignore_case = TRUE))
  }
  
  # ---- Text filters ----
  if (!is.null(name_search))       results <- results %>% filter(safe_text_filter(name, name_search))
  if (!is.null(state_search))      results <- results %>% filter(safe_text_filter(state, state_search))
  if (!is.null(location_search))   results <- results %>% filter(safe_text_filter(location, location_search))
  if (!is.null(coordinates_search))results <- results %>% filter(safe_text_filter(coordinates, coordinates_search))
  if (!is.null(automated_search))  results <- results %>% filter(safe_text_filter(automated, automated_search))
  if (!is.null(lens_search))       results <- results %>% filter(safe_text_filter(current_lens, lens_search))
  
  # ---- Numeric filters ----
  if (!is.null(min_year))   results <- results %>% filter(!is.na(year_first_lit) & year_first_lit >= min_year)
  if (!is.null(max_year))   results <- results %>% filter(!is.na(year_first_lit) & year_first_lit <= max_year)
  if (!is.null(min_height)) results <- results %>% filter(!is.na(height_ft) & height_ft >= min_height)
  if (!is.null(max_height)) results <- results %>% filter(!is.na(height_ft) & height_ft <= max_height)
  
  # ---- Return results ----
  if (nrow(results) == 0) {
    message("No lighthouses found matching these criteria.")
    return(NULL)
  } else {
    return(results)
  }
}

# ---- Example Usage ----
# All lighthouses taller than 150 ft
search_lighthouses(lighthouses_clean, min_height = 150)

# Built before 1850 in Massachusetts
search_lighthouses(lighthouses_clean, state_search = "Massachusetts", max_year = 1849)

# Lighthouses with Fresnel lenses
search_lighthouses(lighthouses_clean, lens_search = "Fresnel")

# Specific partial name
search_lighthouses(lighthouses_clean, name_search = "Cape")

# Multiple filters: Massachusetts, Fresnel lens, height > 100
search_lighthouses(lighthouses_clean,
                   state_search = "Massachusetts",
                   lens_search = "Fresnel",
                   min_height = 100)
```

### Coastal Chronicals: Charting Two Centuries of Light

Mapping U.S. lighthouses by year of construction reveals both geographic and historical patterns. The oldest lighthouses are heavily clustered along the Northeast coast, reflecting early colonial settlement, dense shipping activity, and the establishment of major ports such as Boston and New York. Mid-19th-century constructions spread southward along the Atlantic coast, while the West Coast shows a later surge in the late 19th and early 20th centuries, consistent with westward expansion and the growth of Pacific maritime commerce.

```{r map_lighthouses_year, fig.width=10, fig.height=7}
# 1. Clean year: take last 4-digit number as 'Current' tower, handle missing years
lighthouse_map_data <- lighthouses_clean %>%
  mutate(
    year_numbers = str_extract_all(year_first_lit, "\\d{4}"),
    year_num = map_dbl(year_numbers,
                   ~ if(length(.x) == 0) NA_real_
                   else as.numeric(min(.x)))
  ) %>%
  # Keep only rows with valid year and coordinates
  filter(!is.na(lat) & !is.na(lon) & !is.na(year_num)) %>%
  
  # 2. Create historical era bins for stronger visual contrast
  mutate(
    era = case_when(
      year_num < 1800              ~ "Pre-1800",
      year_num >= 1800 & year_num < 1850 ~ "1800–1849",
      year_num >= 1850 & year_num < 1900 ~ "1850–1899",
      year_num >= 1900 & year_num < 1950 ~ "1900–1949",
      year_num >= 1950              ~ "1950+"
    ),
    era = factor(era, levels = c("Pre-1800","1800–1849","1850–1899","1900–1949","1950+"))
  ) %>%
  
  # 3. Reorder so earliest era is drawn last (appears on top)
  arrange(factor(era, levels = c("1800–1849","1850–1899","1900–1949","1950+","Pre-1800")))

# ---- 4. Get U.S. map data ----
us_map <- ggplot2::map_data("state")

# 5. Full Contiguous U.S. map: filter to contiguous bounds
lighthouse_contiguous <- lighthouse_map_data %>%
  filter(lon >= -125 & lon <= -66,
         lat >= 24  & lat <= 50)

# 6. ---- Display ----
theme_set(theme_minimal(base_size = 14))  # consistent theme

ggplot() +
  # U.S. states as base map
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray92", color = "white", linewidth = 0.2) +
  
  # Lighthouse points (solid circles, no outline)
  geom_point(data = lighthouse_contiguous,
             aes(x = lon, y = lat, color = era),
             size = 3.2, alpha = 0.9) +
  
  # Discrete color scale emphasizes early historical clusters
  scale_color_viridis_d(option = "plasma", direction = -1, name = "Era Built") +
  
  labs(title = "U.S. Lighthouses by Era of Construction (Contiguous States)",
       x = "Longitude", y = "Latitude") +
  
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 11)
  )

# 7. Northeast / Mid-Atlantic zoom: focus on oldest lighthouses
lighthouse_northeast <- lighthouse_map_data %>%
  filter(lon >= -80 & lon <= -66,   # Roughly from mid-Atlantic to Maine
         lat >= 36 & lat <= 45)

# 8. ---- Display ----
ggplot() +
  # U.S. states as base map
  geom_polygon(data = us_map, aes(x = long, y = lat, group = group),
               fill = "gray92", color = "white", linewidth = 0.2) +
  
  # Lighthouse points (same ordering ensures Pre-1800 visible)
  geom_point(data = lighthouse_northeast,
             aes(x = lon, y = lat, color = era),
             size = 3.5, alpha = 0.95) +
  
  # Same era color scale for direct comparison
  scale_color_viridis_d(option = "plasma", direction = -1, name = "Era Built") +
  
  labs(title = "Northeast U.S. Lighthouses by Era of Construction",
       x = "Longitude", y = "Latitude") +
  
  coord_quickmap(xlim = c(-80, -66), ylim = c(36, 45)) +
  
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 11)
  )
```

### Boxplots Ahoy: Regional Construction Trends

Faceted boxplots of construction year by state reveal regional variation in lighthouse development. Northeastern states such as Massachusetts, New York, and Rhode Island show early construction peaks in the 18th and early 19th centuries, reflecting dense maritime commerce and colonial-era port activity, while Maine and Delaware are slightly later. Southern states and those along the Gulf Coast generally have later median construction years. West Coast states, including California, Oregon, and Washington, have their earliest lighthouses built in the mid-to-late 19th century, reflecting later settlement and port development.

```{r map_lighthouses_year_alternative, fig.width=10, fig.height=7}
# ---- 1. Choose states manually ----
selected_states <- c("Massachusetts", "Rhode Island", "Maine", "Delaware", "New York", "North Carolina", "Florida", "California", "Oregon", "Washington")

# ---- 2. Prepare data ----
lighthouse_boxplot_selected <- lighthouses_clean %>%
  filter(state %in% selected_states) %>%
  mutate(
    year_first_lit = as.numeric(year_first_lit),
    
    # categorize by era
    era = case_when(
      year_first_lit < 1800                          ~ "Pre-1800",
      year_first_lit >= 1800 & year_first_lit < 1850 ~ "1800–1849",
      year_first_lit >= 1850 & year_first_lit < 1900 ~ "1850–1899",
      year_first_lit >= 1900 & year_first_lit < 1950 ~ "1900–1949",
      year_first_lit >= 1950                         ~ "1950+"
    ),
    era = factor(era, levels = c("Pre-1800","1800–1849","1850–1899","1900–1949","1950+"))
  ) %>%
  filter(!is.na(era)) %>%
  
  # Reorder states by median construction year
  group_by(state) %>%
  mutate(median_year = median(year_first_lit, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(state = factor(state, levels = unique(state[order(median_year)])))

# ---- 3. Plot ----
ggplot(lighthouse_boxplot_selected, aes(x = state, y = year_first_lit)) +
  geom_boxplot(outlier.shape = NA, fill = "gray90", color = "black", alpha = 0.6) +
  geom_jitter(aes(color = era), width = 0.2, size = 2.5, alpha = 0.85) +
  scale_color_viridis_d(option = "plasma", direction = -1, name = "Era Built") +
  coord_flip() +
  labs(
    title = "Distribution of Lighthouse Construction Years for Selected States",
    x = "State",
    y = "Year Built"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 11)
  )
```

### Towers Through Time: How Lighthouses Grew

Plotting lighthouse height against year of construction reveals patterns in architectural and technological evolution alongside wide variation in tower size. Heights show no consistent upward trend; both historical and modern outliers appear. Some early lighthouses were exceptionally tall to guard especially hazardous stretches of coastline. For example, the Cape Hatteras Lighthouse (North Carolina, 187 ft, 1870) was constructed to protect mariners navigating what is now known as the “Graveyard of the Atlantic,” a treacherous offshore region where hundreds - if not thousands - of shipwrecks have occurred. Equipped with a powerful first-order Fresnel lens, the tower maximized visibility along one of the most dangerous corridors on the eastern seaboard.

Conversely, some smaller or more specialized lighthouses, such as the Craighill Channel Upper Range Front Light (Maryland, 15 ft, 1886), were engineered for close-range navigational precision. One of a paired set of range lights marking the second section of the shipping channel into Baltimore Harbor, it functioned not as a long-distance coastal beacon but as a fixed alignment aid guiding vessels safely into a busy port. Tower height ultimately reflected functional needs, geographic context, and technological change: improved materials and lighting systems enabled taller structures, while advances in optical design sometimes reduced the need for large towers.

```{r lighthouse_height_vs_year, fig.width=10, fig.height=7}
# ---- 1. Prepare data ----
lighthouse_height_data <- lighthouses_clean %>%
  # Convert year_first_lit to numeric; non-numeric values become NA
  mutate(year_first_lit = as.numeric(year_first_lit)) %>%
  filter(!is.na(height_ft) & !is.na(year_first_lit)) %>%
  mutate(
    era = case_when(
      year_first_lit < 1800                ~ "Pre-1800",
      year_first_lit >= 1800 & year_first_lit < 1850 ~ "1800–1849",
      year_first_lit >= 1850 & year_first_lit < 1900 ~ "1850–1899",
      year_first_lit >= 1900 & year_first_lit < 1950 ~ "1900–1949",
      year_first_lit >= 1950               ~ "1950+"
    ),
    era = factor(era, levels = c("Pre-1800","1800–1849","1850–1899","1900–1949","1950+"))
  ) %>%
  # Draw earliest era last so bright yellow points appear on top
  arrange(desc(era == "Pre-1800"))

# ---- 2. Display ----
ggplot(lighthouse_height_data, aes(x = year_first_lit, y = height_ft, color = era)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +  # linear regression
  # geom_smooth(method = "loess", se = FALSE, color = "black", linetype = "dashed") + # optional loess instead

  scale_color_viridis_d(option = "plasma", direction = -1, name = "Era Built") +

  # Add small padding to prevent clipping
  scale_x_continuous(
    limits = c(min(lighthouse_height_data$year_first_lit) - 5,
               max(lighthouse_height_data$year_first_lit) + 5)
  ) +
  scale_y_continuous(
    limits = c(0, max(lighthouse_height_data$height_ft) + 20)
  ) +

  labs(
    title = "Lighthouse Height vs. Year of Construction",
    x = "Year Built",
    y = "Height (ft)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 11)
  )
```

### Guided by Light: Technology Meets Architecture

Lens type and automation status are inconsistently reported, limiting detailed quantitative analysis, but the dataset still illustrates historical and technological patterns. Early lighthouses, particularly those from the 18th and early 19th centuries, often used classical Fresnel lenses - such as the first‑order rotating lenses installed in major ports - and were manually operated, reflecting the engineering standards and operational norms of the time. These towers were sited for maximum visibility in busy shipping areas or at heritage locations. Across the dataset, these historical designs coexist with later adaptations, showing how functional requirements, construction methods, and geographic context influenced lighthouse design.

Newer lighthouses increasingly feature automated systems and modern optics. Many modern lenses, such as the VRB‑25, are Fresnel-derived but smaller, lighter, and designed for efficiency rather than sheer range. Automation reduced the need for on-site personnel and allowed construction in smaller or remote locations. The gradual adoption of classical Fresnel lenses in the 19th century and their modern derivatives demonstrates a clear technological evolution that improved the reach and efficiency of coastal lights. Considering age, height, and lens type together, we can trace how design choices, operational practices, and innovations shaped U.S. lighthouse construction over more than two centuries.

```{r lighthouse_RDS_file}
# Create RDS file
dir.create("data", showWarnings = FALSE)

saveRDS(lighthouses_clean, "data/lighthouses_clean.rds")
```