---
title: "Linguistic Ancestry: Hierarchical Diversity in Human Speech"
name: Verity Elliott
output: html_document
---

Human language is not merely a tool for communication but a living map of ancestral migration, cultural divergence, and social consolidation. Much like the biological lineages of ancient trees, languages belong to expansive families that share common roots, branching out over millennia into distinct dialects and standard tongues. Today, the global linguistic landscape is dominated by a small number of "super-families" - large-scale groupings that account for the vast majority of the world's population. By examining languages with at least 50 million native speakers, we can observe the unequal distribution of linguistic influence and the structural relationships that connect seemingly disparate cultures across the globe.

```{r load-packages, warning = FALSE, message=FALSE}
# Load all required packages quietly
suppressMessages({
  library(tidyverse)
  library(rvest)
  library(janitor)
  library(plotly)
})
```

The dataset for this analysis is compiled from the Wikipedia entry "List of languages by number of native speakers," specifically targeting the table of languages with at least 50 million first - language speakers. To prepare this information for hierarchical visualization, several cleaning steps are required to address the complexities of modern web formatting. Raw character strings are stripped of citation artifacts and hidden CSS code used for screen - reader accessibility, which otherwise introduce noise into the linguistic labels. Reported speaker totals are extracted as numeric values, allowing for quantitative aggregation across different levels of the hierarchy.

To facilitate a nested visualization, the cleaned dataset is reshaped into a parent - child relational structure. This involves creating four distinct layers of data: a global root, language families, intermediate branches, and individual languages. Each observation is assigned a unique identifier and linked to a parent category, ensuring that the total population "rolls up" correctly from specific languages to their broader ancestral groups. This transformation moves the data from a simple descriptive list to a consistent topological map, where every linguistic "leaf" is mathematically anchored to its origin.

```{r retrieve_data}
# ---- 1. Scrape table ----
url <- "https://en.wikipedia.org/wiki/List_of_languages_by_number_of_native_speakers"

lang_raw <- read_html(url) %>%
  html_node(xpath = "//table[contains(., 'at least 50 million')]") %>%
  html_table(fill = TRUE) %>%
  clean_names()

# ---- 2. Clean and reshape table ----
lang_final <- lang_raw %>%
  mutate(
    # 1. Remove CSS noise
    across(where(is.character), ~str_remove_all(., "\\.mw-parser-output.*\\}")),
    
    # 2. Remove "N/a" strings
    across(where(is.character), ~str_remove_all(., "(?i)—N/a|N/a")),
    
    # 3. Handle the speakers column
    speakers_num = as.numeric(str_extract(native_speakers_millions, "[0-9.]+")),
    
    # 4. Standard cleaning
    across(where(is.character), ~str_remove_all(., "\\[.*?\\]")),
    across(where(is.character), str_trim)
  ) %>%
  # 5. Fill empty branches with the family name so the hierarchy doesn't break
  mutate(branch = ifelse(branch == "" | is.na(branch), "Other", branch)) %>%
  filter(!is.na(speakers_num))

lang_final <- lang_final %>%
  select(language, language_family, branch, speakers_num)

# ---- 3. Build Parent-Child "sunburst" dataframe (column-cleaned) ----
# Level 1: The Root
root <- data.frame(
  ids = "World", 
  labels = "World", 
  parents = "", 
  values = sum(lang_final$speakers_num)
)

# Level 2: Families (Parent = World)
families <- lang_final %>%
  group_by(language_family) %>%
  summarise(values = sum(speakers_num), .groups = "drop") %>%
  mutate(ids = language_family, 
         labels = language_family, 
         parents = "World") %>%
  select(ids, labels, parents, values) # Keep only the 4 essentials

# Level 3: Branches (Parent = Family)
branches <- lang_final %>%
  group_by(language_family, branch) %>%
  summarise(values = sum(speakers_num), .groups = "drop") %>%
  mutate(ids = paste0(language_family, "-", branch),
         labels = branch,
         parents = language_family) %>%
  select(ids, labels, parents, values)

# Level 4: Languages (Parent = Branch)
languages <- lang_final %>%
  mutate(ids = language,
         labels = language,
         parents = paste0(language_family, "-", branch),
         values = speakers_num) %>%
  select(ids, labels, parents, values)

# Combine into master hierarchy
sunburst_df <- bind_rows(root, families, branches, languages)
```

### Mapping the Global Voice: Structure and Context of the Sunburst

To visualize the interconnected nature of human speech, this analysis utilizes an interactive Sunburst Chart. Unlike a traditional bar graph, which treats data points as isolated units, a sunburst represents data as a nested topology where the proximity and size of slices indicate structural relationships. At the center lies the global total, which radiates outward into the primary language families. Each subsequent ring represents a deeper level of specificity: the secondary ring shows sub - branches - such as the Romance or Indo - Aryan branches of the Indo - European family - and the outermost ring displays individual languages proportional to their number of native speakers.

The figure is designed with radial text orientation to ensure that labels follow the "spoke" of each slice, maintaining a clear directional flow toward the center even as slices become more granular. Color inheritance is used to denote lineage; for instance, every language within the Sino - Tibetan family shares a common color palette, allowing for immediate visual grouping without the need for an external legend. This radial view reveals a stark reality of modern linguistics: a handful of families occupy the vast majority of the chart’s area, reflecting the historical and geopolitical forces that have shaped global communication. By clicking on specific segments, users can "drill down" into the data, effectively zooming in on a single linguistic branch to explore its internal diversity.

```{r sunburst, fig.width=10, fig.height=8}
plot_ly(
  data = sunburst_df,
  ids = ~ids,
  labels = ~labels,
  parents = ~parents,
  values = ~values,
  type = 'sunburst',
  branchvalues = 'total',
  hoverinfo = "label+value+percent entry",
  insidetextorientation = 'radial',
  marker = list(colorscale = "Viridis")
) %>%
  layout(
    title = list(text = "<b>The Architecture of Human Speech</b><br>Languages with >50M Speakers", y = 0.95),
    margin = list(l = 0, r = 0, b = 0, t = 80)
  )
```